Follow-up Samples

### 2017-03-24 submission/manuscript_outline_20161024.Rmd
latest revised draft

# Adenoma Model 
adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)

# SRN Model
srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T)
srn_imp_vars <- read.csv('../data/process/tables/srn_rf_otu_tax.csv', header = T, row.names = 1)
srn_AUC_run_summary <- read.csv('../data/process/tables/srn_Reduced_ROC_model_summary.csv', header = T)

# CRC Model
crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
crc_imp_vars <- read.csv('../data/process/tables/crc_rf_otu_tax.csv', header = T, row.names = 1)
crc_AUC_run_summary <- read.csv('../data/process/tables/crc_Reduced_ROC_model_summary.csv', header = T)

# All model information
all_probs_paired_wilcox <- read.csv('../data/process/tables/all_crc_srn_adn_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
all_pred_stat_summary <- read.csv('../data/process/tables/all_crc_srn_adn_models_summary_info.csv', header = T, stringsAsFactors = F)
all_confu_matrix <- read.csv('../data/process/tables/all_crc_srn_adn_models_confusion_summary.csv', header = T, stringsAsFactors = F)

@@

**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma or SRN cases *(P-value )*.  But there was a trend for increasingly less community similarity between initial and follow up samples from adenoma to SRN to carcinoma.  Despite this, no difference was found in the relative abundance of any OTU before and after treatment for adenoma, SRN, or carcinoma groups *(P-value )*.  Using Random Forest models to assess whether changes in follow up samples were towards a normal community showed that only those with carcinoma had a significant decrease in positive probability *(P-value)*; providing further evidence that treatment has the greatest effect in those with carcinoma. The adenoma model used a total of *model n* OTUs, the SRN model used a total of *model n* OTUs, and the carcinoma model used a total of *model n* OTUs.  A total of *common n* OTUs were common to both models with many classifying to commensal bacteria (e.g. *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).  Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma *(Pvalue )*

@@

***The Bacterial Community:*** 
Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma *(n = )*, SRN *(n = )*, or carcinoma *(n = )* had any broad differences between their initial and follow up samples.  We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma *(P-value 0.001)* [Figure 1A].  Although no significant differences were observed between SRN and carcinoma there was an increase in the dissimilarity of the initial and follow up samples from adenoma *(mean and SD)* to SRN *(mean and SD)* to carcinoma *(mean and SD)* [Figure 1A].  The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] *(PERMANOVA  0.05)*, SRN [FIgure 1C] *(PERMANOVA 0.05)*, and carcinoma [Figure 1C] *(PERMANOVA  0.05)*.  Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity *(PERMANOVA  0.05)*.  There was no difference between initial and follow up samples when investigating alpha diversity metrics for adenoma, SRN, or carcinoma for any metric tested [Table S1].  Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for adenoma, SRN, or carcinoma only [Figure S1].

***Adenoma Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2A].  There was a total of *blank* OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A]. There was a significant difference between the actual and model predicted group stratification calls *(P-value)*.  There was no significant decrease in the positive probability of adenoma between initial and follow up samples *(P_value)* [Figure 2A].

***SRN Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2B].  There was a total of *blank* OTUs in the SRN model [Figure S3B].  Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal. Also similar to the adenoma model there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.  There was no significant decrease in the positive probability of SRN between initial and follow up samples *(P_value)* [Figure 2B].

***Carcinoma Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2C].  Interestingly, the AUCs improved from adenoma to SRN to carcinoma [Figure 2].  There was a total of *blank* OTUs in the carinoma model [Figure S3C].  Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* appear to be important for carcinoma classification [Figure S3C].  Also similar to the adenoma and SRN models there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.  There was a significant decrease in the positive probability of carcinoma between initial and follow up samples *(P_value)* [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.  The one indivdiual still positive for carcinoma after treatment had an increase in carcinoma positive proability on follow up [Figure 2C].

***Adenoma, SRN, and Carcinoma Common OTUs***
We next wanted to know what predictors within the adenoma, SRN, and carcinoma models were similar to each other.  The main purpose was to identify which OTUs could be important at all three stages of disease.  When we compared the three different models with each other there were a total of *number* common OTUs.  Some of the most common taxonomic identifications belonged to *one*, *two*, *three*, *four*, *five*, *six*, and *seven*.  These along with the vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].

***Treatment Affects on Community***
After observing these these changes from treatment we assessed whether chemotherapy or radiation, in the carcinoma group, and surgery, in the adenoma group, impacted the observed results.  In the carcinoma group neither chemotherapy or radiation provided any additive change from initial sample over surgery alone [Table S3]. For the adenoma group there was no difference between those that received surgery and those that did not [Table S4]. For surgery adenoma and SRN were combined due to the low number of surgery occurances.  There was no difference in the proportion of those receiving surgery between the adenoma and SRN groups *(P-value)*.  This data suggests that microbiome changes observed in the carcinoma group were mostly a result from surgery and not from chemotherapy or radiation.

@@

Many of the common OTUs that we identified taxnomically classified to potential butyrate producers (e.g. *Clostridiales*, *Roseburia*,  and *Anaerostipes*)[Table S2].  Other OTUs classified to bacteria that are inhibited by polyphenols (e.g. *Bacteroides*). Both butyrate and polyphenols are thought to be protective against cancer, in part by reducing inflammation [@louis_gut_2014].  These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes. One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or SRN versus carcinoma *(uncorrected P-value 0.05)*, with time passed being less for adenoma *( days)* and SRN *( days)* than carcinoma *( days)*.  These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time. Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1C]. However, there are two reasons that this may not be the case.  First, the SRN group did not have a significant difference and the collection time of their follow up sample was less than the adenoma group.  Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].

>>>>

# Adenoma Model 
adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)
adn_general_summary <- read.csv('../data/process/tables/adn_reduced_auc_summary.csv', header = T, row.names = 1)

# SRN Model
srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T)
srn_imp_vars <- read.csv('../data/process/tables/srn_rf_otu_tax.csv', header = T, row.names = 1)
srn_AUC_run_summary <- read.csv('../data/process/tables/srn_Reduced_ROC_model_summary.csv', header = T)
srn_general_summary <- read.csv('../data/process/tables/srn_reduced_auc_summary.csv', header = T, row.names = 1)

# CRC Model
crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
crc_imp_vars <- read.csv('../data/process/tables/crc_rf_otu_tax.csv', header = T, row.names = 1)
crc_AUC_run_summary <- read.csv('../data/process/tables/crc_Reduced_ROC_model_summary.csv', header = T)
crc_general_summary <- read.csv('../data/process/tables/crc_reduced_auc_summary.csv', header = T, row.names = 1)

# All model information
all_probs_paired_wilcox <- read.csv('../data/process/tables/all_crc_srn_adn_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F, row.names = 4)
all_pred_stat_summary <- read.csv('../data/process/tables/all_crc_srn_adn_models_summary_info.csv', header = T, stringsAsFactors = F) %>% rename(comparison = X)
all_confu_matrix <- read.csv('../data/process/tables/all_crc_srn_adn_models_confusion_summary.csv', header = T, stringsAsFactors = F)

@@

**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05) but not carcinoma versus SRN cases (P-value `r ifelse(change_summary["thetayc", "Pvalue_srnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05).  But there was a trend for increasingly less community similarity between initial and follow up samples from adenoma to SRN to carcinoma.  Despite this, no difference was found in the relative abundance of any OTU before and after treatment for adenoma, SRN, or carcinoma groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).  Using Random Forest models to assess whether changes in follow up samples were towards a normal community showed that only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma. The adenoma model used a total of `r length(rownames(adn_imp_vars))` OTUs, the SRN model used a total of `r length(rownames(srn_imp_vars))` OTUs, and the carcinoma model used a total of `r length(rownames(crc_imp_vars))` OTUs.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g. *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).  Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (Pvalue `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

@@

***The Bacterial Community:*** 
Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma (n = `r change_summary["thetayc", "adn_n"]`), SRN (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between their initial and follow up samples.  We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].  Although no significant differences were observed between SRN and carcinoma there was an increase in the dissimilarity of the initial and follow up samples from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to SRN (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].  The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), SRN [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).  Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was no difference between initial and follow up samples when investigating alpha diversity metrics for adenoma, SRN, or carcinoma for any metric tested [Table S1].  Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for adenoma, SRN, or carcinoma only [Figure S1].

***Adenoma Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification havin an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].  There was a total of `r length(rownames(adn_imp_vars))` OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A]. There was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was no significant decrease in the positive probability of adenoma between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].

***SRN Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].  There was a total of `r length(rownames(srn_imp_vars))` OTUs in the SRN model [Figure S3B].  Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal. Also similar to the adenoma model there was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was no significant decrease in the positive probability of SRN between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].

***Carcinoma Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].  Interestingly, the AUCs improved from adenoma to SRN to carcinoma [Figure 2].  There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carinoma model [Figure S3C].  Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* appear to be important for carcinoma classification [Figure S3C].  Also similar to the adenoma and SRN models there was a significant difference between the actual and model predicted group stratification calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was a significant decrease in the positive probability of carcinoma between initial and follow up samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.  The one indivdiual still positive for carcinoma after treatment had an increase in carcinoma positive proability on follow up [Figure 2C].

***Adenoma, SRN, and Carcinoma Common OTUs***
We next wanted to know what predictors within the adenoma, SRN, and carcinoma models were similar to each other.  The main purpose was to identify which OTUs could be important at all three stages of disease.  When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs.  Some of the most common taxonomic identifications belonged to *Bacteroides*, *Blautia*, *Anaerostipes*, *Lachnospiraceae*, and *Dorea*.  These along with the vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].

***Treatment Affects on Community***
After observing these these changes from treatment we assessed whether chemotherapy or radiation, in the carcinoma group, and surgery, in the adenoma group, impacted the observed results.  In the carcinoma group neither chemotherapy or radiation provided any additive change from initial sample over surgery alone (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S3]. For the adenoma group there was a single difference in observed OTUs (sobs) between those that received surgery and those that did not (P-value `r ifelse(min(comb_adn_surg_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S4]. For the surgery comparison, adenoma and SRN were combined due to the low number of surgery occurances in these two groups.  There was no difference in the proportion of those receiving surgery between the adenoma and SRN groups (P-value `r ifelse(comb_adn_surg_summary["prop_surg_adn_srn", "pvalue"] < 0.05, invisible("<"), invisible(">"))` 0.05).  This data suggests that microbiome changes observed in the carcinoma group were mostly a result from surgery and not from chemotherapy or radiation.

@@

Many of the common OTUs that we identified taxnomically classified to potential butyrate producers (e.g. *Clostridiales*, *Roseburia*,  and *Anaerostipes*) [Table S2].  Other OTUs classified to bacteria that are inhibited by polyphenols (e.g. *Bacteroides*). Both butyrate and polyphenols are thought to be protective against cancer, in part by reducing inflammation [@louis_gut_2014].  These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes. One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or SRN versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and SRN (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).  These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time. Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1]. However, there are two reasons that this may not be the case.  First, the SRN group did not have a significant difference and the collection time of their follow up sample was less than the adenoma group.  Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].
